Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-06-26 paper/preprint/paper.tex
postponing appendix

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
between the components can governed through penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

@@

@@

@@

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

  \begin{tabular}{cccccccc}
    \toprule
    \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
    Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
    CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
    Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
    Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
    GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
    LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
    MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
    Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
    OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\bottomrule
  \end{tabular}

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Latin Modern Roman}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and, (2) the adversarial loss,
\(\text{advloss}(.)\), with respect to nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
between the components can governed through penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

\begin{figure*}

@@

\end{figure*}%

@@

\begin{table*}

@@

\end{table*}%

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-default}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{table*}
  \begin{tabular}{cccccccc}
    \toprule
    \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
    Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
    CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
    Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
    Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
    GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
    LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
    MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
    Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
    OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\bottomrule
  \end{tabular}
\end{table*}

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/cali/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}

\caption{\label{fig-tune_lr-plaus-cali}Average outcomes for the
plausibility measure across key hyperparameters. This shows the \%
change from the baseline model for the distance-based implausibility
metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
across evaluation runs and test settings (varying parameters for
\emph{ECCo}). Data: California Housing.}
